\section{Distance, Velocity, Acceleration}\label{sec:IntVel}
We recall a general principle that will later be applied to
distance-velocity-acceleration problems, among other things.  If $F(u)$ is
an anti-derivative of $f(u)$, then $\ds \int_a^bf(u)\,du=F(b)-F(a)$.  Suppose
that we want to let the upper limit of integration vary, i.e., we replace
$b$ by some variable $x$.  We think of $a$ as a fixed starting value $x_0$.
In this new notation the last equation (after adding $F(a)$ to both sides)
becomes: 
$$
  F(x)=F(x_0)+\int_{x_0}^xf(u)\,du.
$$ 
Here $u$ is the variable of
integration, called a ``dummy variable,'' since it is not the variable in
the function $F(x)$.  In general, it is not a good idea to use the same
letter as a variable of integration and as a limit of integration.  That
is, $\ds \int_{x_0}^xf(x)dx$ is bad notation, and can lead to errors and
confusion.

An important application of this principle occurs when we are
interested in the position of an object at time $t$ (say, on the
$x$-axis) and we know its position at time $\ds t_0$.  Let $s(t)$ denote
the position of the object at time $t$ (its distance from a reference
point, such as the origin on the $x$-axis).  Then the net change in
position between $\ds t_0$ and $t$ is $\ds s(t)-s(t_0)$.  Since $s(t)$ is an
anti-derivative of the velocity function $v(t)$, we can write
$$
  s(t)=s(t_0)+\int_{t_0}^tv(u)du.
$$
Similarly, since the velocity is an anti-derivative of the acceleration
function $a(t)$, we have 
$$
  v(t)=v(t_0)+\int_{t_0}^ta(u)du.
$$

\begin{example}{Constant Force}{ConstantForce}
Suppose an object is acted upon by a constant
force $F$.  Find $v(t)$ and $s(t)$.
\end{example}

\begin{solution} 
By Newton's law $F=ma$, so the acceleration is
$F/m$, where $m$ is the mass of the object.  Then we first have
$$
  v(t)=v(t_0)+\int_{t_0}^t{F\over m}\,du=v_0+
  \left.{F\over m}u\right|_{t_0}^t=v_0+{F\over m}(t-t_0),
$$ 
using the usual convention $\ds v_0=v(t_0)$.
Then
\begin{eqnarray*}
  s(t)&=s(t_0)+\int_{t_0}^t\left(v_0+{F\over m}(u-t_0)\right)du=s_0+
  \left.(v_0u+{F\over2m}(u-t_0)^2)\right|_{t_0}^t\cr
      &=s_0+v_0(t-t_0)+{F\over2m}(t-t_0)^2.\cr
\end{eqnarray*}
For instance, when $F/m=-g$ is the constant of gravitational acceleration,
then this is the falling body formula (if we neglect air resistance)
familiar from elementary physics:
$$s_0+v_0(t-t_0)-{g\over2}(t-t_0)^2,$$
or in the common case that $\ds t_0=0$,
$$s_0+v_0t-{g\over2}t^2.$$
\end{solution}

Recall that the integral of the velocity function gives the \ifont{net} distance
traveled. If you want to know the \ifont{total} distance traveled, 
you must find out where the velocity function
crosses the $t$-axis, integrate separately over the time intervals when
$v(t)$ is positive and when $v(t)$ is negative, and add up the absolute
values of the different integrals.  For example, if an object is thrown
straight upward at 19.6 m/sec, its velocity function is
$v(t)=-9.8t+19.6$, using $g=9.8$ m/sec for the force of gravity.
This is a straight line which is positive for $t<2$ and negative for $t>2$.
The net distance traveled in the first 4 seconds is thus
$$\int_0^4(-9.8t+19.6)dt=0,$$
 while the total distance traveled in the first
4 seconds is
$$
  \int_0^2(-9.8t+19.6)dt+\left|\int_2^4(-9.8t+19.6)dt\right|=19.6+|-19.6|=39.2
$$ 
meters, $19.6$ meters up and $19.6$ meters down.

\begin{example}{Net and Total Distance}{NetTotalDistance}
The acceleration of an object is given by $a(t)=\cos(\pi
t)$, and its velocity at time $t=0$ is $1/(2\pi)$.  Find both the net and the
total distance traveled in the first 1.5 seconds.
\end{example}

\begin{solution} 
We compute 
$$
  v(t)=v(0)+\int_0^t\cos(\pi u)du={1\over 2\pi}+\left.{1\over\pi}
  \sin(\pi u)\right|_0^t={1\over\pi}\bigl({1\over2}+\sin(\pi t)\bigr).
$$
The \ifont{net} distance traveled is then
\begin{eqnarray*}
  s(3/2)-s(0)&=&\int_0^{3/2}{1\over\pi}\left({1\over2}+\sin(\pi t)\right)\,dt\cr
  &=&\left.{1\over\pi}\left({t\over2}-{1\over\pi}\cos(\pi t)\right)
  \right|_0^{3/2}\cr
  &=&{3\over4\pi}+{1\over\pi^2}\approx 0.340 \hbox{ meters.}\cr
\end{eqnarray*}
To find the {\it total} distance traveled, we need to know when
$(0.5+\sin(\pi t))$ is positive and when it is negative.  This
function is 0 when $\sin(\pi t)$ is $-0.5$, i.e., when $\pi t=7\pi/6$,
$11\pi/6$, etc.  The value $\pi t=7\pi/6$, i.e., $t=7/6$, is the only
value in the range $0\le t\le 1.5$.  Since $v(t)>0$ for $t<7/6$ and
$v(t)<0$ for $t>7/6$, the total distance traveled is
$$ \int_0^{7/6}\ds{1\over \pi}\left({1\over2}+\sin(\pi t)\right)\,dt+
  \Bigl|\int_{7/6}^{3/2} 
  {1\over \pi}\left({1\over2}+\sin(\pi t)\right)\,dt\Bigr|
$$
\begin{eqnarray*}
%& \ds \int_0^{7/6}\ds{1\over \pi}\left({1\over2}+\sin(\pi t)\right)\,dt+
%  \Bigl|\int_{7/6}^{3/2} 
%  {1\over \pi}\left({1\over2}+\sin(\pi t)\right)\,dt\Bigr|\\
 ~\qquad~&=&\ds{1\over \pi}\left( {7\over 12}+{1\over \pi}\cos(7\pi/6)+{1\over
    \pi}\right)+
  {1\over \pi}\Bigl|{3\over 4}-{7\over 12}
  +{1\over \pi}\cos(7\pi/6)\Bigr|\\
  &=&\ds{1\over \pi}\left( {7\over 12}+{1\over \pi}{\sqrt3\over2}+{1\over
    \pi}\right)+
  {1\over \pi}\Bigl|{3\over 4}-{7\over 12}
  +{1\over \pi}{\sqrt3\over2}.\Bigr|\\
  &\approx& 0.409 \hbox{ meters.}
\end{eqnarray*}
\end{solution}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Opensolutionfile{solutions}[ex]
\section*{Exercises for Section \ref{sec:IntVel}}

\begin{enumialphparenastyle}

%%%%%%%%%%
\begin{ex}
 An object moves so that its velocity at time $t$ is
$v(t)=-9.8t+20$ m/s. Describe the motion of the object between $t=0$ and
$t=5$, find the total distance traveled by the object during that
time, and find the net distance traveled.
\begin{sol}
 It rises until $t=100/49$, then falls. The position of the
object at time $t$ is $\ds s(t)=-4.9t^2+20t+k$. The net distance traveled
is $-45/2$, that is, it ends up $45/2$ meters below where it started.
The total distance traveled is $6205/98$ meters. 
\end{sol}
\end{ex}

%%%%%%%%%%
\begin{ex}
 An object moves so that its velocity at time $t$ is $v(t)=\sin t$.
Set up and evaluate a single definite integral to compute the
net distance traveled between $t=0$ and $t=2\pi$.
\begin{sol}
 $\ds\int_0^{2\pi}\sin t\,dt=0$
\end{sol}
\end{ex}

%%%%%%%%%%
\begin{ex}
 An object moves so that its velocity at time $t$ is
$v(t)=1+2\sin t$ m/s. Find the net distance traveled by the object
between $t=0$ and $t=2\pi$, and find the total distance traveled
during the same period.
\begin{sol}
 net: $2\pi$, total: $\ds 2\pi/3+4\sqrt3$ 
\end{sol}
\end{ex}

%%%%%%%%%%
\begin{ex}
 Consider the function $f(x)=(x+2)(x+1)(x-1)(x-2)$ on
$[-2,2]$. Find the total area between the curve and the $x$-axis
(measuring all area as positive).
\begin{sol}
 $8$
\end{sol}
\end{ex}

%%%%%%%%%%
\begin{ex}
 Consider the function $\ds f(x)=x^2-3x+2$ on
$[0,4]$. Find the total area between the curve and the $x$-axis
(measuring all area as positive).
\begin{sol}
 $17/3$
\end{sol}
\end{ex}

%%%%%%%%%%
\begin{ex}
 Evaluate the three integrals:
$$
  A=\int_0^3 (-x^2+9)\,dx\qquad B=\int_0^{4} (-x^2+9)\,dx\qquad 
  C=\int_{4}^3 (-x^2+9)\,dx,
$$
and verify that $A=B+C$.
\begin{sol}
 $A=18$, $B=44/3$, $C=10/3$
\end{sol}
\end{ex}

\end{enumialphparenastyle}