\section{Infinite Limits and Limits at Infinity}\label{sec:InfLimits}
We occasionally want to know what happens to some quantity
when a variable gets very large or ``goes to infinity''.

\begin{example}{Limit at Infinity}{LimitInfinity}
What happens to the function $\ds \cos(1/x)$ as $x$ goes to infinity? It
seems clear that as $x$ gets larger and larger, $1/x$ gets closer and
closer to zero, so $\cos(1/x)$ should be getting closer and closer to 
$\cos(0)=1$.
\end{example}

As with ordinary limits, this concept of ``limit at infinity'' can be
made precise. Roughly, we want $\ds \lim_{x\to \infty}f(x)=L$ to mean that
we can make $f(x)$ as close as we want to $L$ by making $x$ large
enough.

\begin{definition}{Limit at Infinity (Formal Definition)}{LimitAtInfinity}
If $f$ is a function, we say that $\ds \lim_{x\to
  \infty}f(x)=L$ if for every $\epsilon>0$ there is an $N > 0$ so that
  whenever $x>N$, $|f(x)-L|<\epsilon$.
We may similarly define $\ds \lim_{x\to-\infty}f(x)=L$.
\end{definition}

We include this definition for completeness, but we will not explore it in
detail. Suffice it to say that such limits behave in much the same way
that ordinary limits do; in particular there is a direct analog of 
Theorem~\ref{thm:PropertiesLimits}.

\begin{example}{Limit at Infinity}{LimitInfinity2}
Compute $\ds\lim_{x\to \infty}{2x^2-3x+7\over x^2+47x+1}$.
\end{example}

\begin{solution} 
As $x$ goes to infinity both the numerator and denominator go to
infinity. We divide the numerator and
denominator by $\ds x^2$:
$$\lim_{x\to \infty}{2x^2-3x+7\over x^2+47x+1}=
\lim_{x\to \infty}{2-\ds{3\over x}+\ds{7\over x^2}\over
1+\ds{47\over x}+\ds{1\over x^2}}.$$
Now as $x$ approaches infinity, all the quotients with some power of
$x$ in the denominator approach zero, leaving 2 in the numerator and 1
in the denominator, so the limit again is 2.
\end{solution}

In the previous example, we \ifont{divided by the highest power of $x$ that occurs in the denominator} in order to evaluate the limit.
We illustrate another technique similar to this.

\begin{example}{Limit at Infinity}{LimitInfinity3}
Compute the following limit:
$$\lim_{x\to\infty}\frac{2x^2+3}{5x^2+x}.$$
\vspace{-0.5cm}
\end{example}

\begin{solution} 
As $x$ becomes large, both the numerator and denominator become large, so it isn't clear what happens to their ratio. The highest power of $x$ in the denominator is $x^2$, therefore we will divide every term in both the numerator and denominator by $x^2$ as follows:
$$\lim_{x\to\infty}\frac{2x^2+3}{5x^2+x}=\lim_{x\to\infty}\frac{2+3/x^2}{5+1/x}.$$
Most of the limit rules from last lecture also apply to infinite limits, so we can write this as:
$$=\frac{\ds{\lim_{x\to\infty}2+3\lim_{x\to\infty}\frac{1}{x^2}}}{\ds{\lim_{x\to\infty}5+\lim_{x\to\infty}\frac{1}{x}}}=\frac{2+3(0)}{5+0}=\frac{2}{5}.$$
Note that we used the theorem above to get that $\ds{\lim_{x\to\infty}\frac{1}{x}=0}$ and $\ds{\lim_{x\to\infty}\frac{1}{x^2}=0}$.

A shortcut technique is to analyze only the \ifont{leading terms} of the numerator and denominator. A leading term is a term that has the highest power of $x$. If there are multiple terms with the same exponent, you must include all of them.\\
\ifont{Top:} The leading term is $2x^2$.\\
\ifont{Bottom:} The leading term is $5x^2$.\\
Now only looking at leading terms and ignoring the other terms we get:
$$\lim_{x\to\infty}\frac{2x^2+3}{5x^2+x}=\lim_{x\to\infty}\frac{2x^2}{5x^2}=\frac{2}{5}.$$
\end{solution}

We next look at limits whose value is infinity (or minus infinity).

\begin{definition}{Infinite Limit (Useable Definition)}{Infinite Limit}
In general, we will write
$$\lim_{x\to a}f(x)=\infty$$
if we can make the value of $f(x)$ arbitrarily large by taking $x$ to be sufficiently close to $a$ (on either side of $a$) but not equal to $a$.
Similarly, we will write
$$\lim_{x\to a}f(x)=-\infty$$
if we can make the value of $f(x)$ arbitrarily large and \blue{negative} by taking $x$ to be sufficiently close to $a$ (on either side of $a$) but not equal to $a$.
\end{definition}

This definition can be modified for one-sided limits as well as limits with $x\to a$ replaced by $x\to\infty$ or $x\to-\infty$.

\begin{example}{Limit at Infinity}{LimitInfinity4}
Compute the following limit: $\ds\lim_{x\to\infty}(x^3-x).$
\end{example}

\begin{solution} 
One might be tempted to write:
$$\lim_{x\to\infty}x^3-\lim_{x\to\infty}x=\infty-\infty,$$
however, we do not know what $\infty-\infty$ is, as $\infty$ is not a real number and so cannot be treated like one.
We instead write:
$$\lim_{x\to\infty}(x^3-x)=\lim_{x\to\infty}x(x^2-1).$$
As $x$ becomes arbitrarily large, then both $x$ and $x^2-1$ become arbitrarily large, and hence their product $x(x^2-1)$ will also become arbitrarily large. Thus we see that
$$\lim_{x\to\infty}(x^3-x)=\infty.$$
\end{solution}

\begin{example}{Limit at Infinity and Basic Functions}{LimitInfinity5}
We can easily evaluate the following limits by observation:
\[ \begin{array}{ll}
1.~\ds\lim_{x\to \infty} \frac{6}{\sqrt{x^3}}=0 & \qquad 2.~\ds\lim_{x\to -\infty} x-x^2=-\infty \\
\\
3.~\ds\lim_{x\to \infty}x^3+x=\infty            & \qquad 4.~\ds\lim_{x\to \infty} \cos(x)=\mbox{DNE} \\
\\
5.~\ds\lim_{x\to \infty} e^x=\infty             & \qquad 6.~\ds\lim_{x\to -\infty} e^x=0 \\
\\
7.~\ds\lim_{x\to 0^+} \ln x=-\infty             & \qquad 8.~\ds\lim_{x\to 0} \cos(1/x)=\mbox{DNE} \\
\end{array} \]
\end{example}

Often, the shorthand notation $\ds\frac{1}{0^+}=+\infty$ and $\ds\frac{1}{0^-}=-\infty$ is used to represent the following two limits respectively:
$$\lim_{x\to 0^+}\frac{1}{x}=+\infty\qquad\mbox{and}\qquad\lim_{x\to 0^-}\frac{1}{x}=-\infty.$$
Using the above convention we can compute the following limits.

\begin{example}{Limit at Infinity and Basic Functions}{LimitInfinity6}
Compute $\ds{\lim_{x\to 0^+} e^{1/x}}$, $\ds{\lim_{x\to 0^-} e^{1/x}}$ and $\ds{\lim_{x\to 0} e^{1/x}}$.
\end{example}

\begin{solution} 
We have:
$$\lim_{x\to 0^+} e^{\frac{1}{x}}=e^{\frac{1}{0^+}}=e^{+\infty}=\infty.$$
$$\lim_{x\to 0^-} e^{\frac{1}{x}}=e^{\frac{1}{0^-}}=e^{-\infty}=0.$$
Thus, as left-hand limit $\neq$ right-hand limit, 
$$\lim_{x\to 0} e^{\frac{1}{x}}=\mbox{DNE}.$$
\end{solution}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Subsections to include
\input{3-limits/3-5-1-vertical-asymptotes}
\input{3-limits/3-5-2-horizontal-asymptotes}
\input{3-limits/3-5-3-slant-asymptotes}
\input{3-limits/3-5-4-end-behaviour-growth-rate}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Opensolutionfile{solutions}[ex]
\section*{Exercises for \ref{sec:InfLimits}}

\begin{enumialphparenastyle}

%%%%%%%%%%
\begin{ex}
Compute the following limits.
\begin{multicols}{3}
\begin{enumerate}
	\item	$\ds\lim_{x\to \infty} \sqrt{x^2+x}-\sqrt{x^2-x}$
	\item	$\ds\lim_{x\to\infty} {e^x + e^{-x}\over e^x -e^{-x}}$
	\item	$\ds\lim_{t\to1^+}{(1/t)-1\over t^2-2t+1}$
	\item	$\ds\lim_{t\to\infty}{t+5-2/t-1/t^3\over 3t+12-1/t^2}$
	\item	$\ds\lim_{y\to\infty}{\sqrt{y+1}+\sqrt{y-1}\over y}$
	\item	$\ds\lim_{x\to 0^+}{3+x^{-1/2}+x^{-1}\over 2+4x^{-1/2}}$
	\item	$\ds\lim_{x\to\infty}{x+x^{1/2}+x^{1/3}\over x^{2/3}+x^{1/4}}$
	\item	$\ds\lim_{t\to\infty}{1-\sqrt{t\over t+1}\over 2-\sqrt{4t+1\over t+2}}$
	\item	$\ds\lim_{t\to\infty}{1-{t\over t-1}\over 1-\sqrt{t\over t-1}}$
	\item	$\ds\lim_{x\to-\infty}{x+x^{-1}\over 1+\sqrt{1-x}}$
	\item	$\ds\lim_{x\to1^+}{\sqrt{x}\over x-1}$
	\item	$\ds\lim_{x\to\infty}{x^{-1}+x^{-1/2}\over x+x^{-1/2}}$
	\item	$\ds\lim_{x\to\infty}{x+x^{-2}\over 2x+x^{-2}}$
	\item	$\ds\lim_{x\to\infty}{5+x^{-1}\over 1+2x^{-1}}$
	\item	$\ds\lim_{x\to\infty}{4x\over\sqrt{2x^2+1}}$
	\item	$\ds\lim_{x\to\infty}{(x+5)\left({1\over 2x}+{1\over x+2}\right)}$
	\item	$\ds\lim_{x\to0^+}{(x+5)\left({1\over 2x}+{1\over x+2}\right)}$
	\item	$\ds\lim_{x\to2}{x^3-6x-2\over x^3-4x}$
\end{enumerate}
\end{multicols}
\begin{sol}
\begin{multicols}{3}
\begin{enumerate}
	\item	$1$
	\item	$1$
	\item	$-\infty$
	\item	$1/3$
	\item	$0$
	\item	$\infty$
	\item	$\infty$
	\item	$2/7$
	\item	$2$
	\item	$-\infty$
	\item	$\infty$
	\item	$0$
	\item	$1/2$
	\item	$5$
	\item	$\ds 2\sqrt2$
	\item	$3/2$
	\item	$\infty$
	\item	does not exist
\end{enumerate}
\end{multicols}
\end{sol}
\end{ex}


%%%%%%%%%%
\begin{ex} 
The function $\ds f(x) = {x\over\sqrt{x^2+1}}$ has two horizontal asymptotes.  Find them and give a rough sketch of $f$ with its horizontal asymptotes. 
\begin{sol}
$y=1$ and $y=-1$
\end{sol}
\end{ex}

%%%%%%%%%%
\begin{ex}
Find the vertical asymptotes of $\ds f(x)=\frac{\ln x}{x-2}$.
\begin{sol}
	$x=0$ and $x=2$.
\end{sol}
\end{ex}

%%%%%%%%%%
\begin{ex}
Suppose that a falling object reaches velocity $v(t)=50(1-e^{-t/5})$ at time $t$, where distance is measured in $m$ and time $s$. What is the object's terminal velocity, i.e. the value of $v(t)$ as $t$ goes to infinity?
\end{ex}

%%%%%%%%%%
\begin{ex}
Find the slant asymptote of $f(x)=\dfrac{x^2+x+6}{x-3}$.
\begin{sol}
	$y=x+4$
\end{sol}
\end{ex}


%%%%%%%%%%
\begin{ex}
	Compute the following limits.
	\begin{enumerate}
		\item	$\ds\lim_{x\to -\infty}(2x^3-x)$
		\item	$\ds\lim_{x\to \infty}\tan^{-1}(e^x)$
		\item	$\ds\lim_{x\to -\infty}\tan^{-1}(e^x)$
		\item	$\ds\lim_{x\to \infty}\dfrac{e^x+x^4}{x^3+5\ln x}$
		\item	$\ds\lim_{x\to \infty}\dfrac{2^x+5(3^x)}{3(2^x)-3^x}$
		\item	$\ds\lim_{x\to -\infty}\dfrac{2^x+5(3^x)}{3(2^x)-3^x}$
		\item	$\ds\lim_{x\to 0^{+}}\sqrt{x}\ln x$ [Hint: Let $t=1/x$]
	\end{enumerate}
	\begin{sol}
		\begin{enumerate}
			\item	$-\infty$
			\item	$\pi/2$
			\item	0
			\item	$\infty$
			\item	$-5$
			\item	1/3
			\item	0
		\end{enumerate}
	\end{sol}
\end{ex}


\end{enumialphparenastyle}