\section{Analytic Geometry}\label{sec:AnalyticGeometry}

\input{1-review/1-2-new-cartesian-coordinate-plane} 


Suppose we have two points $A$ and $B$ in the $(x,y)$-plane.
We often want to know the change in $x$-coordinate (also called the
``horizontal distance'') in going from $A$ to $B$.  This is often
written $\Delta x$, where the meaning of $\Delta$ (a capital delta in
the Greek alphabet) is ``change in''. Thus, $\Delta x$ can be read as
``change in $x$'' although it usually is read as ``delta $x$''. The
point is that $\Delta x$ denotes a single number, and should not be
interpreted as ``delta times $x$''. Similarly, the ``change in $y$'' is written $\Delta y$
and represents the difference between the $y$-coordinates of the 
two points. It is the vertical distance you have to move in going from $A$ to $B$.

\begin{example}{Change in $x$ and $y$}{ChangeIn}\label{ChangeIn}
	If $A=(2,1)$ and $B=(3,3)$ the change in $x$ is
	$$\Delta x=3-2=1$$
	while the change in $y$ is
	$$\Delta y= 3-1=2$$
	\vspace{-0.5cm}
\end{example}

The general formulas for the change in $x$ and the change in $y$ 
between a point $(x_1,y_1)$ and a point $(x_2,y_2)$ are:
$$
\Delta x=x_2-x_1,\qquad\qquad\Delta y=y_2-y_1.
$$
Note that either or both of these might be negative.\\

%%%%%%%%%% APPLICATION
%%%%%%%%%%%%%%%%%%%%%
%In applications, often letters other than $x$ and $y$ are used, and
%often different scales are chosen in the horizontal and vertical
%directions.
%
%\pagebreak 
%\begin{example}{Data Plot}{DataPlot}
%	Suppose you drop a coin from a window, and you want to study how 
%	its height above the ground changes from second to second.
%	It is natural to let the letter $t$ denote the time
%	(the number of seconds since the object was released) and to let the
%	letter $h$ denote the height.  For each $t$ (say, at one-second
%	intervals) you have a corresponding height $h$.  This information can
%	be tabulated, and then plotted on the $(t,h)$ coordinate plane, as
%	shown in Figure~\ref{fig:data plot}.
%\end{example}
%\begin{solution}
%
%\hspace{1cm}	\begin{tabular}{c|c}
%	Time, $t$ & Height, $h$  \\ \hline
%	0 & 80 \\
%	1 & 75.1 \\
%	2 & 60.4 \\
%	3 & 35.9 \\
%	4 & 1.6 \\ 
%	\end{tabular} 
%	
%	\vspace{-2cm}
%	
%	\figure[h]	
%		\hspace{70mm}	\beginpicture
%			\normalgraphs
%			%\ninepoint
%			\setcoordinatesystem units <2truecm,0.5truemm>
%			\setplotarea x from 0 to 4.2, y from 0 to 90
%			\axis left shiftedto x=0 ticks numbered from 20 to 80 by 20 /
%			\axis bottom shiftedto y=0 ticks numbered from 0 to 4 by 1 /
%		%	\put {$t$} [l] <3pt,0pt> at 4.2 0
%		%	\put {$h$} [l] <3pt,0pt> at 0 90
%			\setquadratic
%			\plot 0 80 1 75.1 2 60.4 3 35.9 4 1.6 /
%			\multiput {$\bullet$} at 0 80 1 75.1 2 60.4 3 35.9 4 1.6 /
%			\endpicture 
%		
%
%
%\vspace{-4.4cm} \hspace{63mm} \rotatebox{90}{\textcolor{black}{Height, $h$ (meters)}}
%
%\vspace{8.6mm} \hspace{95mm} Time, $t$(seconds) \\
%
%\caption{A data plot, height versus time. \label{fig:data plot}}
%\endfigure
%\end{solution}




% Input each subsection
\input{1-review/1-2-new-relations}
\input{1-review/1-2-1-lines}
\input{1-review/1-2-2-distances-and-midpoints}
\input{1-review/1-2-3-conics}
		

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Opensolutionfile{solutions}[ex]
\section*{Exercises for \ref{sec:AnalyticGeometry}}

\begin{enumialphparenastyle}

\begin{multicols}{2}
%%%%%%%%%%
\begin{ex}
Find the equation of the line in the form $y=mx+b$:
\begin{enumerate}
	\item	through $(1,1)$ and $(-5, -3)$
	\item	through $(-1,2)$ with slope $-2$
	\item	through $(-1,1)$ and $(5, -3)$
	\item	through $(2,5)$ and parallel to the line $3x+9y+6=0$
	\item	with $x$-intercept 5 and perpendicular to the line $y=2x+4$
\end{enumerate}
\begin{sol}
\begin{enumerate}
	\item	$(2/3)x+(1/3)$
	\item	$y=-2x$
	\item	$y=(-2/3)x+(1/3)$
	\item	$y=-x/3+17/3$
	\item	$y=-1/2x+5/2$
\end{enumerate}
\end{sol}
\end{ex}

%%%%%%%%%%
\begin{ex}
Change the following equations to the form $y=mx+b$, graph the
line, and find the $y$-intercept and $x$-intercept.
\begin{enumerate}
	\item	$y-2x=2$
	\item	$x+y=6$
	\item	$x=2y-1$
	\item	$3=2y$
	\item	$2x+3y+6=0$
\end{enumerate}
\begin{sol}
\begin{enumerate}
	\item	$y=2x+2$, $2$, $-1$
	\item	$y=-x+6$, $6$, $6$
	\item	$y=x/2+1/2$, $1/2$, $-1$
	\item	$y=3/2$, $y$-intercept: $3/2$, no $x$-intercept
	\item	$y=(-2/3)x-2$, $-2$, $-3$
\end{enumerate}
\end{sol}
\end{ex}
 
%%%%%%%%%%
\begin{ex}
Determine whether the lines $3x+6y=7$ and $2x+4y=5$ are parallel.
\begin{sol}
Yes, the lines are parallel as they have the same slope of $-1/2$
\end{sol}
\end{ex}

%%%%%%%%%%
\begin{ex}
Suppose a triangle in the $(x,y)$-plane has vertices $(-1,0)$,
$(1,0)$ and $(0,2)$.  Find the equations of the three lines that lie along
 the sides of the triangle in $y=mx+b$ form.
\begin{sol}
$y=0$, $y=-2x+2$, $y=2x+2$
\end{sol}
\end{ex}

%%%%%%%%%%
\begin{ex}
Let $x$ stand for temperature in degrees Celsius (centigrade), and let
$y$ stand for temperature in degrees Fahrenheit.  A temperature of $0^\circ$C
corresponds to $32^\circ$F, and a temperature of
$100^\circ$C corresponds to $212^\circ$F.  Find the
equation of the line that relates temperature Fahrenheit $y$ to
temperature Celsius $x$ in the form $y=mx+b$.  
Graph the line, and find the point at which this line intersects $y=x$.
What is the practical meaning of this point?
\begin{sol}
$y=(9/5)x+32$, $(-40,-40)$
\end{sol}
\end{ex}

%%%%%%%%%%
\begin{ex}
A car rental firm has the following charges for a certain type of car:
\$25 per day with 100 free miles included, \$0.15 per mile for more than
100 miles.  Suppose you want to rent a car for one day, and you know you'll
use it for more than 100 miles.  What is the equation relating the cost
$y$ to the number of miles $x$ that you drive the car?
\begin{sol}
$y=0.15x+10$
\end{sol}
\end{ex}

%%%%%%%%%%
\begin{ex}
A photocopy store advertises the following prices: 5c~per
copy for the first 20 copies, 4c~per copy for the 21st through
100th copy, and 3c~per copy after the 100th copy.  Let $x$ be the
number of copies, and let $y$ be the total cost of photocopying.  (a)
Graph the cost as $x$ goes from 0 to 200 copies.  (b) Find the
equation in the form $y=mx+b$ that tells you the cost of making $x$
copies when $x$ is more than 100.
\begin{sol}
$0.03x+1.2$
\end{sol}
\end{ex}

%%%%%%%%%%
\begin{ex}
Market research tells you that if you set the price of an item at
\$1.50, you will be able to sell 5000 items; and for every 10 cents you
lower the price below \$1.50 you will be able to sell another 1000 items.
Let $x$ be the number of items you can sell, and let $P$ be the price of
an item.  (a)~Express $P$ linearly in terms of $x$, in other words,
express $P$ in the form $P=mx+b$.  (b)~Express $x$
linearly in terms of $P$.
\begin{sol}
(a) $P=-0.0001x+2$\\
(b) $x=-10000P+20000$
\end{sol}
\end{ex}

%%%%%%%%%%
\begin{ex}
An instructor gives a 100-point final exam, and decides that a score
90 or above will be a grade of 4.0, a score of 40 or below will be a grade
of 0.0, and between 40 and 90 the grading will be linear.  Let $x$ be
the exam score, and let $y$ be the corresponding grade.  Find a formula
of the form $y=mx+b$ which applies to scores $x$ between 40 and 90.
\begin{sol}
$(2/25)x-(16/5)$
\end{sol}
\end{ex}

%%%%%%%%%%%
%\begin{ex}
%Find the distance between the pairs of points:
%\begin{enumerate}
%	\item	$(-1,1)$ and $(1,1)$.
%	\item	$(5,3)$ and $(-7,-2)$.
%	\item	$(1,1)$ and the origin.
%\end{enumerate}
%\begin{sol}
%\begin{enumerate}
%	\item	$2$
%	\item	$\sqrt{2}$
%	\item	$\sqrt{2}$
%\end{enumerate}
%\end{sol}
%\end{ex}
%
%%%%%%%%%%%
%\begin{ex}
%Find the midpoint of the line segment joining the point $(20,-10)$ to the origin.
%\end{ex}

%%%%%%%%%%%%%%%%%%%%
\begin{ex}
Find the distance $d$ between the points and the midpoint $M$ of the line segment which connects them.\\
\begin{enumerate} 
 \item \hspace{2mm} $(1,2)$, $(-3,5)$ \\
 \item  \hspace{2mm} $(3, -10)$, $(-1, 2)$ \\ 
 \item 	\hspace{2mm} $\left( \dfrac{1}{2}, 4\right)$, $\left(\dfrac{3}{2}, -1\right)$ \\
 \item  \hspace{2mm}	 $\left(- \dfrac{2}{3}, \dfrac{3}{2} \right)$, $\left(\dfrac{7}{3}, 2\right)$ \\
 \item  \hspace{2mm} $\left( \dfrac{24}{5}, \dfrac{6}{5} \right)$, $\left( -\dfrac{11}{5}, -\dfrac{19}{5} \right)$ \\
 \item  \hspace{2mm}	 $\left(\sqrt{2}, \sqrt{3}\right)$, $\left(-\sqrt{8}, -\sqrt{12}\right)$ \vphantom{$\left( \dfrac{6}{5} \right)$} \\
 \item  \hspace{2mm} $\left(2 \sqrt{45}, \sqrt{12} \right)$, $\left(\sqrt{20}, \sqrt{27} \right)$\\
 \item  \hspace{2mm} $(0, 0)$, $(x, y)$ \\
\end{enumerate}
\begin{sol}
	\begin{enumerate} 
\item $d = 5$, $M = \left(-1, \frac{7}{2} \right)$
\item $d = 4 \sqrt{10}$, $M = \left(1, -4 \right)$
\item $d = \sqrt{26}$, $M = \left(1, \frac{3}{2} \right)$
\item $d= \frac{\sqrt{37}}{2}$, $M = \left(\frac{5}{6}, \frac{7}{4} \right)$
\item  $d = \sqrt{74}$, $M = \left(\frac{13}{10}, -\frac{13}{10} \right)$ \vphantom{$\left( \frac{\sqrt{3}}{2} \right)$}
\item $d= 3\sqrt{5}$, $M = \left(-\frac{\sqrt{2}}{2}, -\frac{\sqrt{3}}{2} \right)$
\item  $d = \sqrt{83}$, $M = \left(4 \sqrt{5}, \frac{5 \sqrt{3}}{2} \right)$
\item $d = \sqrt{x^2 + y^2}$, $M = \left( \frac{x}{2}, \frac{y}{2}\right)$ \vphantom{$\left( \frac{\sqrt{3}}{2} \right)$}
\end{enumerate}
\end{sol}	
\end{ex}

\begin{ex}
Find all of the points of the form $(x, -1)$ which are $4$ units from the point $(3,2)$.
\begin{sol}
$(3 + \sqrt{7}, -1)$, $(3-\sqrt{7}, -1)$	
\end{sol}	
\end{ex}

\begin{ex}
Find all of the points on the $y$-axis which are $5$ units from the point $(-5,3)$.
\begin{sol}
	$(0,3)$
\end{sol}	
\end{ex}

\begin{ex}
Find all of the points on the $x$-axis which are $2$ units from the point $(-1,1)$.
\begin{sol}
$(-1+\sqrt{3},0)$, $(-1-\sqrt{3},0)$ \vphantom{$\left( \frac{\sqrt{3}}{2} \right)$}	
\end{sol}	
\end{ex}

\begin{ex}
Find all of the points of the form $(x,-x)$ which are $1$ unit from the origin.
\begin{sol}
	$\left(\frac{\sqrt{2}}{2},-\frac{\sqrt{2}}{2} \right)$, $\left(-\frac{\sqrt{2}}{2},\frac{\sqrt{2}}{2}\right)$
\end{sol}	
\end{ex}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{ex}
 Let's assume for a moment that we are standing at the origin and the positive $y$-axis points due North while the positive $x$-axis points due East.  Our Sasquatch-o-meter tells us that Sasquatch is 3 km West and 4 km South of our current position.  What are the coordinates of his position?  How far away is he from us?  If he runs 7 km due East what would his new position be?
 \begin{sol}
 	$(-3,-4), \, 5 \,\, km, \,\, (4, -4)$
 \end{sol}
 \end{ex}	
 
 %%%%%%%%%%%%%%%%%%%%%%%%%
\begin{ex}
Verify the Distance Formula  for the cases when:
 
\begin{enumerate}
 	
 	\item The points are arranged vertically.  (Hint: Use $P(a, y_{\mbox{\tiny$0$}})$ and $Q(a, y_{\mbox{\tiny$1$}})$.)
 	\item The points are arranged horizontally. (Hint: Use $P(x_{\mbox{\tiny$0$}}, b)$ and $Q(x_{\mbox{\tiny$1$}}, b)$.)
 	\item The points are actually the same point. 
 	
 \end{enumerate}
\begin{sol}
	
	
\end{sol}
\end{ex}

%%%%%%%%%%%%%%%%%%%
\begin{ex}
Verify the Midpoint Formula by showing the distance between $P(x_{\mbox{\tiny$1$}}, y_{\mbox{\tiny$1$}})$ and $M$ and the distance between $M$ and $Q(x_{\mbox{\tiny$2$}}, y_{\mbox{\tiny$2$}})$ are both half of the distance between $P$ and $Q$. 
\begin{sol}
	
\end{sol}
\end{ex}
	 
%%%%%%%%%%%%%%%%%%%%
\begin{ex}
Show that the points $A$, $\;B$ and $C$ below are the vertices of a right triangle.
\begin{enumerate}
	
	\item  $A(-3,2)$, $\;B(-6,4)$, and $C(1,8)$
	
	\item   $A(-3, 1)$, $\;B(4, 0)$ and $C(0, -3)$
	
	
\end{enumerate}
\begin{sol}
	(a) The distance from $A$ to $B$ is $|AB| = \sqrt{13}$, the distance from $A$ to $C$ is $|AC| = \sqrt{52}$, and the distance from $B$ to $C$ is $|BC| = \sqrt{65}$.  Since $\left(\sqrt{13}\right)^2 + \left( \sqrt{52} \right)^2 = \left( \sqrt{65} \right)^2$, we are guaranteed by the \href{http://en.wikipedia.org/wiki/Pythagorean_theorem#Converse}{\underline{converse of the Pythagorean Theorem}} that the triangle is a right triangle.\\
	(b) Show that $|AC|^{2} + |BC|^{2} = |AB|^{2}$.
\end{sol}
\end{ex}
	 
 
%%%%%%%%%%%%%%%%%%%
\begin{ex}
Find a point $D(x, y)$ such that the points $A(-3, 1)$, $\;B(4, 0)$, $\;C(0, -3)$ and $D$ are the corners of a square.  Justify your answer.
\begin{sol}
	
\end{sol}
\end{ex}		
	


%%%%%%%%%%
\begin{ex}
Find the equation of the circle of radius 3 centered at:
\begin{multicols}{2}
\begin{enumerate}
	\item	$(0,0)$
	\item	$(5,6)$
	\item	$(-5,-6)$
	\item	$(0,3)$
	\item	$(0,-3)$
	\item	$(3,0)$
\end{enumerate}
\end{multicols}
\begin{sol}
\begin{enumerate}
	\item	$x^2+y^2=9$
	\item	$(x-5)^2+(y-6)^2=9$
	\item	$(x+5)^2+(y+6)^2=9$
\end{enumerate}
\end{sol}
\end{ex}

%%%%%%%%%%
\begin{ex}
For each pair of points $A(x_1,y_1)$ and $B(x_2,y_2)$ find an equation of
the circle with center at $A$ that goes through $B$.
\begin{multicols}{2}
\begin{enumerate}
	\item	$A(2,0)$, $B(4,3)$
	\item	$A(-2,3)$, $B(4,3)$
\end{enumerate}
\end{multicols}
\end{ex}

%%%%%%%%%%
\begin{ex}
Determine the type of conic section and sketch it.
\begin{enumerate}
	\item	$x^2+y^2+10y=0$
	\item	$9x^2-90x+y^2+81=0$
	\item	$6x+y^2-8y=0$
\end{enumerate}
\begin{sol}
\begin{enumerate}
	\item	circle
	\item	ellipse
	\item	horizontal parabola
\end{enumerate}
\end{sol}
\end{ex}

%%%%%%%%%%
\begin{ex} 
Find the standard equation of the circle passing through $(-2,1)$
 and tangent to the line $3x-2y =6$ at the point $(4,3)$.  Sketch. 
 (Hint: The line through the center of the circle and the point of tangency
 is perpendicular to the tangent line.)
\begin{sol}
$(x+2/7)^2+(y-41/7)^2=1300/49$
\end{sol}
\end{ex}

\end{multicols}

\end{enumialphparenastyle}