\section{Taylor Series}\label{sec:taylorseries}








































% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
We have seen that some functions can be represented as series, which
may give valuable information about the function. So far, we have seen
only those examples that result from manipulation of our one
fundamental example, the geometric series. We would like to start with
a given function and produce a series to represent it, if possible.

Suppose that $\ds f(x)=\sum_{n=0}^\infty a_nx^n$ on some interval of
convergence centered at 0. Then we know that we can compute derivatives of $f$ by
taking derivatives of the terms of the series. Let's look at the first
few in general:
\begin{align*}
  f'(x)&=\sum_{n=1}^\infty n a_n x^{n-1}=a_1 + 2a_2x+3a_3x^2+4a_4x^3+\cdots	\\
  f''(x)&=\sum_{n=2}^\infty n(n-1) a_n x^{n-2}=2a_2+3\cdot2a_3x
    +4\cdot3a_4x^2+\cdots	\\
  f'''(x)&=\sum_{n=3}^\infty n(n-1)(n-2) a_n x^{n-3}=3\cdot2a_3
    +4\cdot3\cdot2a_4x+\cdots	\\
\end{align*}

By examining these it's not hard to discern the general pattern. The
$k$th derivative must be
\begin{align*}
  f^{(k)}(x)&=\sum_{n=k}^\infty n(n-1)(n-2)\cdots(n-k+1)a_nx^{n-k}	\\
  &=k(k-1)(k-2)\cdots(2)(1)a_k+(k+1)(k)\cdots(2)a_{k+1}x+{}	\\
  &\qquad {}+(k+2)(k+1)\cdots(3)a_{k+2}x^2+\cdots	\\
\end{align*}
We can express this more clearly by using factorial notation:
\[
  f^{(k)}(x)=\sum_{n=k}^\infty {n!\over (n-k)!}a_nx^{n-k}=
  k!a_k+(k+1)!a_{k+1}x+{(k+2)!\over 2!}a_{k+2}x^2+\cdots
\]

We can solve for $a_n$ by substituting $x=0$ in the formula for $f^{(k)}(x)$:
\[f^{(k)}(0)=k!a_k+\sum_{n=k+1}^\infty {n!\over (n-k)!}a_n0^{n-k}=k!a_k,\]
\[a_k={f^{(k)}(0)\over k!}.\]
Note that the original series for $f$ yields $f(0)=a_0$.

So if a function $f$ can be represented by a series, we can easily find such a series.
Given a function $f$, the series
\[\sum_{n=0}^\infty {f^{(n)}(0)\over n!}x^n\]
is called the \dfont{Maclaurin 
series} for $f$.

\begin{example}{Maclaurin Series}{MacSeriesOne}
Find the Maclaurin series for $f(x)=1/(1-x)$.
\end{example}
\begin{solution}
We need to
compute the derivatives of $f$ (and hope to spot a pattern).
\begin{align*}
  f(x)&=(1-x)^{-1}	\\
  f'(x)&=(1-x)^{-2}	\\
  f''(x)&=2(1-x)^{-3}	\\
  f'''(x)&=6(1-x)^{-4}	\\
  f^{(4)}(x)&=4!(1-x)^{-5}	\\
  &\vdots	\\
  f^{(n)}(x)&=n!(1-x)^{-n-1}	\\
\end{align*}

So
\[a_n={f^{(n)}(0)\over n!}={n!(1-0)^{-n-1}\over n!}=1\]
and the Maclaurin series is
\[\sum_{n=0}^\infty 1\cdot x^n=\sum_{n=0}^\infty x^n,\]
the geometric series.
\end{solution}

A warning is in order here. Given a function $f$ we may be able to
compute the Maclaurin series, but that does not mean we have found a
series representation for $f$. We still need to know where the series
converges, and if, where it converges, it converges to $f(x)$. While
for most commonly encountered functions the Maclaurin series does
indeed converge to $f$ on some interval, this is not true of all
functions, so care is required.

As a practical matter, if we are interested in using a series to
approximate a function, we will need some finite number of terms of
the series. Even for functions with messy derivatives we can compute
these using computer software like Sage. If we want to describe a series
completely, we would like to be able to write down a formula for a typical
term in the series. Fortunately, a few of the most important functions are very
easy.

\begin{example}{Maclaurin Series}{MacSeriesTwo}
Find the Maclaurin series for $\sin x$.
\end{example}
\begin{solution}
Computing the first few derivatives is simple: $f'(x)=\cos x$, $f''(x)=-\sin x$,
$f'''(x)=-\cos x$, $\ds f^{(4)}(x)=\sin x$, and then the pattern
repeats. The values of the derivative when $x=0$ are:
1, 0, $-1$, 0, 1, 0, $-1$, 0,\dots, and so the Maclaurin series is
\[
  x-{x^3\over 3!}+{x^5\over 5!}-\cdots=
  \sum_{n=0}^\infty (-1)^n{x^{2n+1}\over (2n+1)!}.
\]

We should always determine the radius of convergence:
\[
  \lim_{n\to\infty} {|x|^{2n+3}\over (2n+3)!}{(2n+1)!\over |x|^{2n+1}}
  =\lim_{n\to\infty} {|x|^2\over (2n+3)(2n+2)}=0,
\]
so the series converges for every $x$. Since it turns out that this
series does indeed converge to $\sin x$ everywhere, we have a series
representation for $\sin x$ for every $x$.
\end{solution}

Sometimes the formula for the $n$th derivative of a function $f$ is
difficult to discover, but a combination of a known Maclaurin series
and some algebraic manipulation leads easily to the Maclaurin series
for $f$.

\begin{example}{Maclaurin Series}{MacSeriesThree}
Find the Maclaurin series for $x\sin(-x)$.
\end{example}
\begin{solution}
To get from $\sin x$ to $x\sin(-x)$ we substitute $-x$ for $x$ and
then multiply by $x$. We can do the same thing to the series for $\sin
x$:
\[
  x\sum_{n=0}^\infty (-1)^n{(-x)^{2n+1}\over (2n+1)!}
  =x\sum_{n=0}^\infty (-1)^{n}(-1)^{2n+1}{x^{2n+1}\over (2n+1)!}
  =\sum_{n=0}^\infty (-1)^{n+1}{x^{2n+2}\over (2n+1)!}.
\]
\end{solution}

As we have seen, a power series can be centered at a point
other than zero, and the method that produces the Maclaurin series can
also produce such series.

\begin{example}{Taylor Series}{TaylorSeriesOne}
Find a series centered at $-2$ for $1/(1-x)$.
\end{example}
\begin{solution}
If the series is $\ds\sum_{n=0}^\infty a_n(x+2)^n$ then looking at the
$k$th derivative:
$$k!(1-x)^{-k-1}=\sum_{n=k}^\infty {n!\over (n-k)!}a_n(x+2)^{n-k}$$
and substituting $x=-2$ we get
$\ds k!3^{-k-1}=k!a_k$ and $\ds a_k=3^{-k-1}=1/3^{k+1}$, so the series is
$$\sum_{n=0}^\infty {(x+2)^n\over 3^{n+1}}.$$
\end{solution}

Such a series is called the 
\dfont{Taylor series} for the function,
and the general term has the form
\[{f^{(n)}(a)\over n!}(x-a)^n.\]

A Maclaurin series is simply a Taylor series with $a=0$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Opensolutionfile{solutions}[ex]
\section*{Exercises for \ref{sec:taylorseries}}

\begin{enumialphparenastyle}

\begin{ex}
For each function, find the Maclaurin series or Taylor series centered
at $a$, and the radius of convergence.

\begin{enumerate}
	\item $\cos x$
	\item $\ds e^x$
	\item $1/x$, $a=5$
	\item $\ln x$, $a=1$
	\item $\ln x$, $a=2$
	\item $\ds 1/x^2$, $a=1$
	\item $\ds 1/\sqrt{1-x}$
	\item Find the first four terms of the Maclaurin series for $\tan
	x$ (up to and including the $\ds x^3$ term).
	\item Use a combination of Maclaurin series and algebraic
	manipulation to find a series centered at zero for
	$\ds x\cos (x^2)$.
	\item Use a combination of Maclaurin series and algebraic
	manipulation to find a series centered at zero for
	$\ds xe^{-x}$.
\end{enumerate}
\begin{sol}
\begin{enumerate}
	\item $\ds\sum_{n=0}^\infty (-1)^n x^{2n}/(2n)!$, $R=\infty$
	\item $\ds\sum_{n=0}^\infty x^n/n!$, $R=\infty$
	\item $\ds\sum_{n=0}^\infty (-1)^n{(x-5)^n\over 5^{n+1}}$, $R=5$
	\item $\ds\sum_{n=1}^\infty (-1)^{n-1}{(x-1)^n\over n}$, $R=1$
	\item $\ds\ln(2)+\sum_{n=1}^\infty (-1)^{n-1}{(x-2)^n\over n 2^n}$, $R=2$
	\item $\ds\sum_{n=0}^\infty (-1)^n(n+1)(x-1)^n$, $R=1$
	\item $\ds1+\sum_{n=1}^\infty {1\cdot3\cdot5\cdots(2n-1)\over
		n!2^n} x^n=1+\sum_{n=1}^\infty {(2n-1)!\over 2^{2n-1}(n-1)!\,n!}x^n$, $R=1$
	\item $\ds x+x^3/3$
	\item $\ds\sum_{n=0}^\infty (-1)^n x^{4n+1}/(2n)!$
	\item $\ds\sum_{n=0}^\infty (-1)^n x^{n+1}/n!$
\end{enumerate}
\end{sol}
\end{ex}

\end{enumialphparenastyle}